---
title: "Making a dashboard accessible"
format: 
  html:
    toc: true
    toc-expand: 1
    toc-location: left
    toc-depth: 3
    page-layout: full
---

## Introduction

This demonstration begins with a draft dashboard called "Fun with  Irises", 
which displays data from the `iris` dataset included with R. 
The purpose of this demonstration is to introduce you to relatively easy to
implement solutions for select accessibility issues that I have encountered 
when designing my own dashboards. 

What this demonstration does not cover:

1. Font and color considerations to make your charts and text easier to read.
1. Using sound and texture to make your content more accessible. 
    For an example of how to implement the R packages
    [`tactileR`](https://github.com/jooyoungseo/tactileR) and 
    [`sonify`](https://cran.r-project.org/package=sonify), 
    see the code for my 
    [texture-sound app](https://github.com/ajstamm/shiny-a11y-app/tree/master/texture-sound).
1. Making sure language, vocabulary, and messaging are appropriate for the 
    intended audience.

## The dashboard

The "Fun with Irises" dashboard includes three tabs and a sidebar. 
The tabs include:

* Introduction. This tab is text only, with no interactive elements.
* Barplot. This tab includes a bar plot that can be modified using 
    the first three filters in the sidebar.
* Histogram. This tab includes a histogram that can be modified using 
    all of the filters in the sidebar.
    
Download the script to run the dashboard yourself at #TODO

## Accessibility review results

There are several free tools that can be used to test for accessibility in Shiny dashboards. We have listed three of the ones we use below.

* [`shinya11y`](https://github.com/ewenme/shinya11y) is an R package 
   for testing dashboard accessibility maintained by @ewenme on GitHub. 
* [Accessible Name and Description Inspector (ANDI)](https://www.ssa.gov/accessibility/andi/help/install.html) 
   is a bookmarklet for testing webpage accessibility maintained by 
   the US Social Security Administration. 
* [NonVisual Desktop Access (NVDA)](https://www.nvaccess.org/download/) is a 
   screen reader maintained by the non-profit NV Access.

By using these tools, and by attempting to navigate the dashboard using a 
keyboard, we have discovered several issues with accessing information.
For this demonstration, we will review the following issues.

Screen reader users cannot access some content.

1. The dashboard's language is not defined.
1. The only way to access the information in the charts is by looking at them.
2. There is no alternative text for charts.
3. Some filter labels are ignored by the screen reader.

Keyboard only users cannot access some content.

5. The two slider filters lack keyboard focus.
6. The double-headed slider filter does not work with the keyboard.

Content is unclear or hard to understand.

7. The dashboard does not have a title.
8. There is only one way to navigate to different parts of the dashboard.
9. There are no navigation instructions.
10. There are no descriptions or explanations for the charts.
11. The filters are labelled only with vague keywords.
12. The filters display for all tabs, even ones where they are not relevant.

This demonstration will cover how to address each of the above issues in turn.

## Solutions to address screen reader issues

### Define the dashboard language

This fix is easy. I have added it to my dashboard template so that unless I am
working on a non-English dashboard, I never need to think about it again.

In the UI settings, add the line, `tags$html(lang = "en")`. 

If you want to add a language other than English, one place you can find the correct tag is by using the [Language subtag lookup app](https://r12a.github.io/app-subtags/).

To check this fix with ANDI, activate ANDI, choose "Structures" from the drop-down, expand the "more details" menu, and select "page title".


### Add a second way to access chart information

My preferred alternate method to display data is to include a table below each 
chart. Basic tables are accessible by screen readers. 
In a production app, I will also include a button to download the data as a 
comma-separated value (CSV) file.

In the server function, I will include a `renderDT` call for each table that 
includes the code to create the same table that is used to generate the chart.
These calls will look similar to the one below. 

```{r ex_table_def}
#| eval: FALSE
output$bar_table <- DT::renderDT({
  d <- filtered_data() |> group_by(Species) |> summarise(Count = n())
  dt <- DT::datatable(d, rownames = FALSE, class = 'cell-border stripe',
                      options = list(paging = FALSE, searching = FALSE))
  return(dt)
})
```

Then after the chart in the appropriate tab panel, I add the code,
`dataTableOutput("bar_table")`.


### Add alternative text to charts

General guidance suggests that alternative (or alt) text should not be more 
than 250 characters. If your alternative text is likely to be longer than that, 
consider moving part of it to an explanatory paragraph with your chart. 
Remember, basic unformatted text is fully accessible to screen readers. 
A general guideline is that all information should be available in a table, 
the alternative text, and an accompanying paragraph, 
but details do not need to be repeated across all three.

For this example, my alternative text includes basic summary information for my 
chart and a note to review the accompanying table for more information. In all, 
my alternative text is around 120 characters long. 
Note that it does not include all of the information in the table. 

```{r ex_bar_alt}
#| eval: false
output$bar_plot <- renderPlot({
  d <- filtered_data() |> group_by(Species) |> dplyr::summarise(Count = n())
  ggplot(data = d, aes(x = Species, y = Count)) +
         geom_bar(stat = "identity", fill = "#003865") +
         labs(title = "Frequency chart of iris species")
  },
  alt = reactive({
    d <- filtered_data()
    aria <- paste("Barplot of", input$species, "irises filtered on", 
                  min(d$measure), "to", max(d$measure), input$measure, ".",
                  "Refer to the table below the chart for values displayed.")
    return(aria)
  }))
```

The `alt` tag above assigns the text that the screen reader will read when it 
reaches the chart.

To check this fix with `shinya11y` and a mouse, click on the spectacles in the lower left, then click "Screen Reader Wand". Hover over the chart and read the alternate text in the wand window.


### Ensure filters are screen reader accessible

In the drop-down filters, the screen reader skips over the labels when reading 
the values. 
To address this, add the option `selectize = FALSE` in the input function. 
Note that this only works if you do not need any of the javascript *selectize* 
features.

```{r ex_selectize}
#| eval: false
selectInput("species", selectize = FALSE, label = "Species:", selected = "All", 
            choices = c("All", unique(as.character(iris$Species))))
```

To check this fix with NVDA and a mouse, activate NVDA then click the top of the sidebar and listen to NVDA read through the content.


## Solutions to address keyboard access issues

### Change the single-headed slider filter to a drop-down

Sliders are problematic when navigating via keyboard. 
When tabbing onto the single-ended slider, the keyboard focus does not show up, 
so the only way to determine the cursor location is to click arrows to try to 
change the bins.
I design my dashboards to use drop-downs or open text fields instead.

Here is a single-ended alternative to the bins slider.

```{r ex_single_end}
#| eval: false
selectInput("bins", selectize = FALSE, label = "Number of bins:", 
            choices = 1:20, selected = 10)
```

### Change the double-headed filter to two text boxes

Sliders are problematic when navigating via keyboard. 
When tabbing onto the double-ended slider, keyboard focus is not visible and 
the individual ends cannot be modified with arrow keys. 
I design my dashboards to use drop-downs or open text fields instead.

Here is a double-ended alternative to the measurements slider. Note that they default to the minimum and maximum values, respectively.
For a production-ready dashboard, I might include checks for reasonable numeric 
values and whether the minimum and maximum values are swapped.

```{r ex_double_end}
#| eval: false
splitLayout(
  textInput("range_min", label = HTML("Minimum<br>", measure, ":"), 
            placeholder = min(i$measure), value = min(i$measure)),
  textInput("range_max", label = HTML("Maximum<br>", measure, ":"), 
            placeholder = max(i$measure), value = max(i$measure))
)
```


## Solutions for issues with understanding

### Add a browser tab title

This fix is easy. I have added it to my dashboard template so that aside from 
changing the dashboard's title, I never need to think about it again.

The `h1()` command in the UI settings sets the title within your application. 
However, the tab in the browser contains only numbers for the title. For users 
relying in tab names to locate content, this is confusing.

To add a tab title, add the line `tags$title("Fun with irises")` in the UI 
settings. 
I prefer to add this line near the app title. 
If the app title is short, I will use the same title in both places. 
If the app title is long, I will use a shortened version for the tab title.

To check this fix with ANDI, activate ANDI, choose "Structures" from the drop-down, expand the "more details" menu, and select "page title".



### Ensure there are at least two navigation methods

The only way to navigate to the different parts of the dashboard is to click on 
the tabs. There are two issues with that. First, not everyone may realize the 
tabs are clickable. Second, someone using a screen reader will have no idea 
what the tabs are or that there are more than one until after the screen reader 
finishes reading the sidebar. 

To address both of these issues, I like to include a brief, persistent table of 
contents at the top of the sidebar with internal links to the tabs.

There are three steps to adding internal links. 

1. Ensure that your tabset panel and each of the tab panels inside the tabset 
    panel all have unique and defined IDs. 
    In the example here, the tabset panel has `id = "my_tabs"` and 
    the barplot panel has `value = "tab_bar"`.

2. Set up a reactive element for each tab. 
    The one for the barplot tab looks like this. Note that 
    `inputID` matches the tabset's ID and `selected` matches the tab's value.

```{r ex_link_def}
#| eval: false
observeEvent(input$link_tab_bar, {
  updateTabsetPanel(session, inputId = "my_tabs", selected = "tab_bar")
})
```

3. In the sidebar, add the link to the reactive and a short explanation of the 
    link. I like to use a list format, which I set up using The HTML tags 
    `<ul>` and `<li>` with the syntax shown below.

```{r ex_link_call}
#| eval: false
tags$ul(
  tags$li(strong(actionLink("link_tab_bar", "Barplot")), "- Bar chart")
)
```



### Add navigation instructions

The table of contents will provide some additional instructions. 
Within each chart tab, I will include some additional information about 
what to find where.

```{r ex_instruc}
#| eval: false
p('Select items in the sidebar to control the iris species, measure, and 
   measure limits to display. If the sidebar is not visible, open it by 
   clicking the chevron (">") or "toggle sidebar" label.')
```



### Add chart explanations

Text explanations for charts are useful for reducing the amount of information 
you need to include in alternative text. 
They are also useful for people who have trouble reading charts. 
Ideally your explanation should include why the chart matters and 
what you want the user to take away from it. 

In this example, the point is for you to learn the code, 
not to create a production-ready dashboard with a specific purpose, 
so the chart description here is not very meaningful.

```{r ex_hist_desc}
#| eval: false
output$hist_desc <- shiny::renderText({
  d <- filtered_data()
  desc <- paste("This histogram shows", input$measure, "ranging from", 
                min(d$measure), "to", max(d$measure), "cm for", input$species, 
                "irises. The median is", median(d$measure), "cm and the mean is",
                round(mean(d$measure), digits = 2), "cm.")
  return(paste("<p>", HTML(desc), "</p>"))
})
```

Then before the chart in the appropriate tab panel, I add the code,
`htmlOutput("hist_desc")`.


### Clearly define all filters

Filters of only one or two words may not be understandable to the user. 
Any time you require user input, ensure your request is clear. 
For example, instead of requesting "Height", consider if one of the following 
would work better.

* Enter your estimated height in inches as a whole number
* Your height without shoes in centimeters
* Your height at your last doctor's visit in feet

In the sidebar filter, "Measure" is a vague, unclear term. 
Let's modify it to clarify what we are asking.

```{r ex_label}
#| eval: false
selectInput("measure", selectize = FALSE,
            label = "Select the measurement to filter on for the chart:", 
            choices = names(iris)[1:4], selected = names(iris)[1])
```


### Display filters only for the relevant tabs

Right now the filters are visible no matter which tab we select. 
However, none of them are relevant for the introduction tab and 
the bins filter is not relevant for the barplot tab. Let's fix that.

To control when a filter is displayed, we use the following code to create 
conditional panels. Note that below, one conditional panel is nested inside 
another. The first panel's content only displays when the selected tab is 
*not* the introduction (indicated with `!=`). The second panel has an 
additional filter, to only display when the histogram tab is selected 
(indicated with `==`). Note that the conditions are strings that use 
slightly different format than R code and include nested quotes.

```{r ex_condition}
#| eval: false
    conditionalPanel(
      condition = "input.my_tabs != 'tab_intro'",
      # display this content and 
      conditionalPanel(
        condition = "input.my_tabs == 'tab_hist'",
        # display this content
      )
    )
```




